Fused regression for network inference in multiple species

### 2015-11-16 paper1.tex
hi

\subsubsection{Equivalent prior}
Our L2 fusion penalty is equivalent to assuming a Gaussian prior with variance proportional to $\frac{1}{\lambda_S}$ on differences in parameters with fusion constraints. Combined with the regularization constraints, this forms a multivariate Gaussian prior on weights in both networks. When using our method on systems with incomplete orthology mapping, this results in different prior probability distributions for those parameters with fusion constraints and those without. We adjust for this by solving for a constant to multiply fused interactions by to equalize the volumes of the prior distributions. The intent of this adjustment is to ensure that constrained interactions are not on average more highly penalized, which may tend to drive their weight towards zero, causing them to be excluded from the network.

@@

@@

\subsection{Optimization}
We use glmnet to compute the regularization path for the
 ridge penalty. Given a range of $\lambda_R$ values, we fit models for each $\lambda_R$. We use cross validation to select the optimal parameter, the $\lambda_R$ value which minimizes the error of prediction on the leave out set.

@@

\subsection{Network similarity and fusion performance}
Using related data for network inference allows us to borrow statistical power from other sources, effectively increasing the sample size and boosting the sensitivity and specificity of learned interactions. The decision of what other sources to draw from is an important one; at one extreme is combining unrelated data, and at the other is resampling from the same experiments. We simulated this spectrum by creating pairs of networks and varying the similarity between networks. The main factors governing similarity between our generated networks is the extent (and accuracy) of the orthology mapping, and the magnitude of variability between analogous interactions. We conducted a series of simulations to assess the effect of increasing orthology coverage on networks. In order to determine the effect of the size of the orthology mapping on network inference performance, we simulated a series of 20 TFs by 200 genes networks at several values of percent orthology coverage.

As expected, when the conserved subgraph is very similar, increasing the weight of the fusion constraint $\lambda_S$ improved network recovery (Figure 2). Networks with higher orthology coverage saw a greater benefit from fusion, although the improvement was not limited to analogous interactions (Figure 4). When the conservation is weak, and even the portions of the network which are conserved have deviated from each other significantly, performance may not continue to improve with by increasing the fusion weight; we simulate this by adding noise to the conserved subgraph (Figure 3). We show that even for networks where the conserved subgraph is weakly conserved, there exists a 'sweet spot' where fusion regression improves network recovery.

One way of assessing the effect of fusion on network inference performance is to compute an ``exchange-rate'' between conditions in each network. Suppose you are interested in determining a regulatory network for species A, a model organism which is experimentally inconvenient. A related species, B, is easier to collect data for, but of less intrinsic interest. Our simulation results suggest that adding conditions from species B to a fused network inference problem has an effect on network A recovery that is similar to adding conditions from A at a reduced rate.

\subsection{Fused regression improves performance on both the constrained and non-constrained parts of the network}
Our method learns networks simultaneously, using predictions about conservation of TF-gene interactions to share information between sources. Our approach is useful for learning networks from similar sources such as different cell-lines from the same species, where there exists a one-to-one mapping of genes, as well as datasets where the networks are expected to be more different than similar, where there exists limited orthology mapping. For the latter case, we are interested in knowing if performance gains from fused regression are limited to those interactions which have fusion constraints. To test this, we used sets of 20 TF by 200 gene synthetic networks, with varying proportions of orthology information about TFs and genes. We varied lamS and computed AUPR on the portion of the network that had fusion constraints, the portion where TF-gene interactions from one network could be mapped to putative analogous interactions in the second network based off of orthology; the unconstrained subnetwork; and the whole network. As expected, performance improved as the fusion penalty weight increased. Interestingly, performance gains were seen even in the portion of the network that is unconstrained by fusion. By constraining part of an underdetermined system, we obtain gains in even the unconstrained part.

\subsection{Adaptive fusion successfully identifies and unfuses 'neofunctionalized' genes}
We recognize that orthology prediction is not a perfect proxy for functional conservation. We implemented an adaptive fusion algorithm that attempts to optimize a nonconvex saturating penalty function on differences between fused interactions (Figures 5-6). Pairs of interactions which are very dissimilar even after fusion, which sit in the flat portion of this penalty function, are effectively ``unfused''. Because our network inference problem is underdetermined, meaning that multiple networks would fit our data equally well, the ``unfusing'' of constraints is convincing evidence for the rejection of a network which uses orthology prediction to predict similarity in interaction weights. ''Unfusing'' suggests neofunctionalization of one or more of the genes involved in this pair of interactions.

@@

\subsection{Integrating datasets using fusion regression}
Although there are many large-scale collaborations which attempt to make protocols as uniform as possible for comparability between datasets generated by different labs, there still exists technical and biologial variability between many experiments attempting to capture the same or similar experimental condition. With the advent of new technology, such as RNAseq and single-cell RNAseq, microarray is no longer the dominant assay for genome-wide expression, but it is important to be able to combine datasets generated using different technologies, lest we lose valuable information. Currently, the most widely used approach to combining datasets for network inference is to learn networks from disparate datasets separately, then rank combine the networks as in Marbach et al \cite{marbach_revealing_2010}. We compare this approach to using fusion regression. We include, along with our B. subtilis dataset, a previously published dataset containing 269 samples covering 104 conditions \cite{nicolas2012condition}. We compare performance when learning the networks separately, learning the networks separately followed by rank combining, and learning the networks simultaneously using fusion regression, and show improvement on network inference using our fused L2 approach.

\subsection{Transcription factor activity estimation integrates into fusion regression approach}
We test a combination of our fused regression approach with a method for estimating transcription factor activities (TFA). Rather than modeling gene expression using transcription factor mRNA abundance, we attempt to fit gene expression as a function of transcription factor activity. We estimate TFA based on known regulatory interactions using network component analysis \cite{liao2003network}. We randomly divide the prior known interactions in half, and use half to learn TFA. The rest we reserve as a gold standard for validation. When we incorporate TFA and repeat our experiment combining B. subtilis datasets, we see a marked improvement in performance.

>>>>

\subsection{Iterative solver}

When the number of genes linked by fusion constraints is relatively small, the augmented design matrix approach we have described has the advantage of being efficiently solvable with a closed form solution. However, in the general case the augmented system of equations may be too large to solve efficiently with standard methods. To address this limitation, we developed an iterative solver that uses coordinate-wise descent to solve for solutions corresponding to a sequence of values of fusion penalty weights; since our fused L2 method uses a convex and differentiable penalty function, this approach converges to a global minimizer. Although less efficient than the augmented design matrix approach we developed, the iterative solver has the advantage of computing a solution path for $\lambda_S$.

On each iteration $t$ the iterative solver computes

\begin{equation}
\text{argmin}_{\beta^S(t) : S_i, S_j \in \text{species}} \displaystyle\sum_{S_i, S_j \in \text{species}} ||X^{S_i}\beta^{S_i}(t) - Y^{S_i}||^2 + \lambda_R||\beta^{S_i}(t)||^2 + \displaystyle \sum_{\beta_{g,k}^{S_i} \approx \beta_{h,l}^{S_j}} \lambda_S(\beta^{S_i}_{g,k}(t) - \beta_{h,l}^{S_j}(t-1)^2)
\end{equation}
Note that this is almost identical to equation 5, but now the network $\beta$ is a function of the iteration number $t$. On each step, the we compute $\beta$s that minimize a penalized cost function where the fusion penalty encourages similarity of a pair of parameters, using the previous iteration's solution. This process is iterated until the estimated $\betas$s converge. Because each iteration reduces the error between $\beta(t)$ and $\beta(t-1)$, and because $\beta(t) = \beta(t-1)$ is the globally optimal solution, this process must eventually converge to the same network as (cite 5). Although we have not produced bounds on the convergence rate, in practice a small number of iterations are necessary.

\subsection{Fusion and regularization path}
The iterative solver requires an initial guess $\beta^{S_i}(0)$ for each species $S_i$. When the initial guess is close to the final solution, a small number of iterations are necessary. Because a small change in $\lambda_S$ will tend to produce a small change in the network, solving for the $\lambda_S+\epsilon$ network given $\beta$s initialized to the $\lambda_S$ network requires a small number of iterations. As a consequence, the entire solution path corresponding to a large number of values of $\lambda_S= \lambda_S_1, \lambda_S_2, ..., \lambda_S_{max}$ can be solved efficiently by using the solution of each network as the initial guess for the next.

We use Tibshirani's cyclical coordinate descent algorithms from the 'glmnet' package \cite{friedman_regularization_2010} to compute a ridge regularization path.

\subsection{Optimization}
Because jointly optimizing $\lambda_R$ and $\lambda_S$ is computationally prohibitive, we opted to optimize the parameters separately. First, we use glmnet to compute the regularization path for the ridge penalty, and given a range of $\lambda_R$ values, we fit models for each $\lambda_R$. We use cross validation to select the optimal parameter, the $\lambda_R$ value which minimizes the error of prediction on the leave out set. Following selection of $\lambda_R$, we search for optimal $\lambda_S$ again using cross validation. Note that both parameters are chosen without reference to the gold standard, which is used in a separate evaluation of network quality.

%\subsection{Equivalent prior}
%Our L2 fusion penalty is equivalent to assuming a Gaussian prior with variance proportional to $\frac{1}{\lambda_S}$ on differences in parameters with fusion constraints. Combined with the regularization constraints, this forms a multivariate Gaussian prior on weights in both networks. When using our method on systems with incomplete orthology mapping, this results in different prior probability distributions for those parameters with fusion constraints and those without. We adjust for this by solving for a constant to multiply fused interactions by to equalize the volumes of the prior distributions. The intent of this adjustment is to ensure that constrained interactions are not on average more highly penalized, which may tend to drive their weight towards zero, causing them to be excluded from the network.

@@

Our adaptive penalty function introduces, in addition to regularization and fusion penalty weights $\lambda_R$ and $\lambda_S$, an unknown parameter $a$. We could employ grid search using cross-validation to searach for the best parameters, but this can be computationally expensive. Moreover, we are primarily interested in using this saturating penalty as a way of testing the hypothesis that conservation in GRNs can be predicted based off of known similarities between genes. Therefore, we implement a way of picking $a$ with a user-selected percentile of differences in interaction weights, which represents the user's hypothesis for the proportion of evolutionary constraints which should be relaxed. A gold standard can also be used for selecting $a$ by comparing differences in weights for truly analogous interactions with those of randomly paired interactions, and selecting the difference in interaction weights at which no further fusion penalty is incurred.

@@

@@

\subsection{Using fused regression to learn related networks}
Using related data for network inference allows us to borrow statistical power from other sources, effectively increasing the sample size and boosting the sensitivity and specificity of learned interactions. We used synthetic data to approximate two related biological processes and show that simultaneously learning the networks describing them using fused L2 regression to share information about related interactions results in an improvement in our ability to recover the true networks. Adding conditions from a related dataset has an effect on network inference which is akin to adding experimental data, without doing additional experiments.

The decision of what other sources to draw from is an important one; at one extreme is combining unrelated data, and at the other is resampling from the same experiments. We simulated this spectrum by creating pairs of networks and varying the similarity between networks. The main factors governing similarity between our generated networks is the extent (and accuracy) of the orthology mapping, and the variability between analogous, or conserved, interactions. We conducted a series of simulations to assess the effect of increasing orthology coverage on networks. In order to determine the effect of the size of the orthology mapping on network inference performance, we simulated a series of 20 TFs by 200 genes networks at several values of percent orthology coverage.

As expected, when the conserved subgraph is very similar, increasing the weight of the fusion constraint $\lambda_S$ improved network recovery (Figure 2). Networks with higher orthology coverage saw a greater benefit from fusion.

When the conservation is weak, and even the portions of the network which are conserved have deviated from each other significantly, performance may not continue to improve by increasing the fusion weight; we simulate this by adding noise to the conserved subgraph (Figure 3). We show that even for networks where the conserved subgraph is weakly conserved, there exists a 'sweet spot' where fusion regression improves network recovery.

\subsection{Fused regression improves performance on both the constrained and non-constrained parts of the network}
Our method learns networks simultaneously, using predictions about conservation of TF-gene interactions to share information between sources. Our approach is useful for learning networks from similar sources such as different cell-lines from the same species, where there exists a one-to-one mapping of genes, as well as datasets where the networks are expected to be more different than similar, and there exists limited orthology mapping. For the latter case, we are interested in knowing if performance gains from fused regression are limited to those interactions which have fusion constraints. To test this, we used sets of 20 TF by 200 gene synthetic networks, with varying proportions of orthology information about TFs and genes. We varied lamS and computed AUPR on the portion of the network that had fusion constraints, the portion where TF-gene interactions from one network could be mapped to putative analogous interactions in the second network based off of orthology; the unconstrained subnetwork; and the whole network (Figure 5). As expected, performance improved as the fusion penalty weight increased. Interestingly, performance gains were seen even in the portion of the network that is unconstrained by fusion. By constraining part of an underdetermined system, we obtain gains in even the unconstrained part.

\subsection{Adaptive fusion successfully identifies and unfuses 'neofunctionalized' genes}
We recognize that orthology prediction is not a perfect proxy for functional conservation. We implemented an adaptive fusion algorithm that attempts to optimize a nonconvex saturating penalty function on differences between fused interactions (Figures 5-6). Pairs of interactions which are very dissimilar even after fusion, which sit in the flat portion of this penalty function, are effectively ``unfused,'' and no further penalty is incurred as differences in interaction weights grow. Because our network inference problem is underdetermined, meaning that multiple networks would fit our data equally well, the ``unfusing'' of constraints is convincing evidence for the rejection of a network which uses orthology to predict similarity in interaction weights. ''Unfusing'' suggests neofunctionalization of one or more of the genes involved in this pair of interactions.

@@

\subsection{Integrating datasets using fusion regression}
Although there are many large-scale collaborations which attempt to make protocols as uniform as possible for comparability between datasets generated by different labs, there still exists technical and biologial variability between many experiments attempting to capture the same or similar experimental condition. With the advent of new technology, such as RNAseq and single-cell RNAseq, microarray is no longer the dominant assay for genome-wide expression, but it is important to be able to combine datasets generated using different technologies, lest we lose valuable information. Currently, the most widely used approach to combining datasets for network inference is to learn networks from disparate datasets separately, then rank combine the networks as in Marbach et al \cite{marbach_revealing_2010}. We compare this approach to using fusion regression. We include, along with our B. subtilis dataset, a previously published dataset containing 269 samples covering 104 conditions, obtained using a different microarray and different strain of B. subtilis \cite{nicolas2012condition}. We compare performance when learning the networks separately, learning the networks separately followed by rank combining, and learning the networks simultaneously using fusion regression, and show improvement on network inference using our fused L2 approach. Importantly, our approach simultaneously learns multiple networks, rather than learning networks separately than combining into one.

\subsection{Transcription factor activity estimation integrates into fusion regression approach}
We test a combination of our fused regression approach with a method for estimating transcription factor activities (TFA). Rather than modeling gene expression using transcription factor mRNA abundance, we attempt to fit gene expression as a function of transcription factor activity. We estimate TFA based on known regulatory interactions using network component analysis \cite{liao2003network}. We randomly divide the prior known interactions in half, and use half to learn TFA. The rest we reserve as a gold standard for validation. When we incorporate TFA and repeat our experiment combining B. subtilis datasets, we see a marked improvement in performance from using TFA. Our gains from using fused regression are preserved and even increased with the use of TFA.

\section{Discussion}
The challenge of learning genome-wide transcriptional regulatory networks became possible with the development of assays such as microarray and RNAseq, and the integration of other data types including ChIP SEQ, have improved the quality of GRNs.

We present a method to simultaneously learn related networks by

We present a method for network inference which allows the principled combination of relevant and related data.

As technologies probing gene expression as well mapping chromatin features mature, the potential to create genome-wide maps of gene interactions becomes more realistic. The ability to combine diverse data types, becomes
